---
title: "Echocardiography Analysis"
author: "Mark E. Pepin"
date: "3/09/2021"
output: 
  html_document: 
    toc: yes
    number_sections: yes
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

This script contains the coding needed to import, clean, visualize, and analyze echocardiographic data.

## Clean raw data from Echocardiography tracings.

The reports (.csv) generated from the Echo software produces a poorly-formatted sheet of information that needs to be parsed to extract phenotypic parameters. This function looks within all sub-directories for the echo reports (".csv"), and compiles them. Once compiled, this table is exported as "**Master_Table.xlsx**" in the same directory where the script is run.

```{r clean.raw}
library(openxlsx)
library(tidyr)
library(plyr)
library(dplyr)
library(tidyverse)
Echo_tS<-Sys.time()
temp = list.files(path = "./", pattern="*\\csv", recursive = TRUE)
Timeline = c() #create an empty timeline vector for echo study dates
Master_Table<-as.data.frame(matrix(ncol =4, nrow = 1))
Master_Names<-c("Echo_Param", "Dates", "Value", "Sample_ID")
colnames(Master_Table)<-Master_Names
TableList<-c()
##########################
for (i in 1:length(temp)) { #do the following  (see below) for every echo analysis .csv file found
# Compile ridiculous table containing all data
Echo_Table<-read.table(paste0("./", temp[i]), header = FALSE, sep = ";", fill = TRUE, dec = ",")
Echo_Table[(nrow(Echo_Table)+1):(nrow(Echo_Table)+100),]<-""
## Save Study date
studyDate<-Echo_Table[which(grepl("Study Date", Echo_Table[,1])),2]
  studyDate<-strptime(as.character(studyDate), format = "%d.%m.%Y")
  studyDate<-as.Date(studyDate)
  # studyDate<-as.POSIXct(format(studyDate, "%Y-%m-%d"), format = "%Y-%m-%d") # reformats the date (easier to sort)
  Timeline[i]<-studyDate # put this date into the timeline vector
## Sample names/rows
sampleRows<-which(grepl("Series Name", Echo_Table[,1])) # extract rows for each sample
sampleNames<-Echo_Table[sampleRows,2] # extract sample names
## Sample-specific Study dates
sampleDatesRows<-which(grepl("Series Date", Echo_Table[,1])) # extract rows for each sample
sampleDates<-Echo_Table[sampleDatesRows,2] # extract sample names
  sampleDates<-strptime(as.character(sampleDates), format = "%d.%m.%Y")
  sampleDates<-as.Date(sampleDates)
sampleIndex<-cbind(sampleNames, sampleDates) #sample-date combination
## Create an echo table for each time point/file
Table<-data.frame(sampleNames)
List<-list()
for (j in 1:length(sampleRows)) { # FOR every sample, grab the echo parameters
    j_table<-Echo_Table[(sampleRows[j]):(sampleRows[j]+72),]
    j_tableM<-j_table %>% filter((grepl("MAutoLV", V1) & grepl("M-Mode", V2)) | (grepl("LV Trace", V1) & grepl("M-Mode", V2)) | (grepl("LV Area", V1) & 
      grepl("M-Mode", V2)))
    j_tableB<-j_table %>% filter((grepl("MAutoLV", V1) & grepl("B-Mode", V2)) | (grepl("LV Trace", V1) & grepl("B-Mode", V2)) | (grepl("LV Area", V1) & 
      grepl("B-Mode", V2)))
    j_tableB$V1<-gsub("(Long).*", "\\1", j_tableB$V1)
    j_tableB$V1<-gsub("(Short).*", "\\1", j_tableB$V1)
    j_tableM$V1<-gsub("(MAutoLV).*", "\\1", j_tableM$V1)
    j_tableBM<-rbind(j_tableM, j_tableB)
    j_tableBM<-j_tableBM %>% unite("Echo_Param", V1:V3, remove = T)
    rownames(j_tableBM)<-j_tableBM$Echo_Param
    colnames(j_tableBM)<-c("Echo_Param", "Units", sampleNames[j])
    j_tableBM<-as.data.frame(j_tableBM)
    j_tableBM<-j_tableBM %>% select(-Units)
    j_tableBM$Dates<-sampleDates[j]
    j_tableBM[,sampleNames[j]]<-as.numeric(gsub(",", ".", j_tableBM[,sampleNames[j]])) #Convert Echo values to numeric "." data.
    List[[j]]<-as.data.frame(j_tableBM) #Convert all echo tables into a list 1:length(sampleNames)
    } # Extracts sample echo data from the master table, assembling into a nicer one...
Full_Table<-List %>% reduce(full_join) #merge all echo tables and remove duplicated Measurements (M-Mode)
Echo_Interim<-gather(Full_Table, "Sample", Value, -Dates, -Echo_Param, na.rm = TRUE) %>% mutate(Sample_ID=Sample) %>% select(-Sample) %>% distinct()
# Add these samples/dates to the "Master Table"
TableList[[i]]<-Echo_Interim
}
Echo.Database_gathered<-TableList %>% reduce(full_join)
Echo.Database_spread<-Echo.Database_gathered %>% spread(key = Echo_Param, value = Value)
write.xlsx(Echo.Database_spread, paste0("Echo_Summary.Table_", Sys.Date(), ".xlsx"))
```

## Import Metadata for Mice

This next step is specific for a given project, and requires information about each mouse (genotype, sex, treatment, etc..). **This must meet the following parameters:**

1.  It must be created using an **"\_Index.xlsx**" ending and placed in the same folder location where this script is run.

2.  It must contain a "Sample_ID" column where each mouse identifier is located. NOTE: this must be identical to the identifiers used to label mice within the Echo reports.

3.  **"Duration"** (this is derived from "**Int_Date**" (Date of the intervention) and "**Study_Date**" (Date of Echo)

```{r Index}
library(openxlsx)
library(dplyr)
library(tibble)
Index_loc = list.files(path = "./", pattern="*\\Index.xlsx")
Index<-read.xlsx(Index_loc)
Index<-Index %>% filter(Sample_ID!="NA") %>% mutate(Sample_ID = as.character(Sample_ID)) #Remove rows with missing Sample_ID (without this, there is no point...)
head(Index)
Annotated_echo<-Index %>% full_join(rownames_to_column(Echo.Database_spread), sto.info, by = ("rowname" = "Sample_ID")) #Merge index to Master Echo table
##Format Index with new rows
Annotated_echo$Int_Date<-as.Date(Annotated_echo$Int_Date, origin="1899-12-30") # Intervention Date (Excel time)
Annotated_echo$DOB<-as.Date(Annotated_echo$DOB, origin="1899-12-30") # Geburtsdatum (Excel time)
Annotated_echo$DOD<-as.Date(Annotated_echo$DOD, origin="1899-12-30") # Tot Datum (Excel time)
Annotated_echo<-Annotated_echo %>% mutate(Age.t0=Int_Date-DOB) # Tot Datum (Excel time)
Annotated_echo<-Annotated_echo %>% mutate(Duration=Dates-Int_Date) #Relative Time (Echo - Intervention)
Annotated_echo$Group<-factor(Annotated_echo$Group, levels = c("WT_Sham", "WT_ORAB", "BAP1_Sham", "BAP1_ORAB"))
#Export compiled table
write.xlsx(Annotated_echo, paste0("Annotated.Echo_Summary.Table_", Sys.Date(), ".xlsx"))
```

## Time-Course Study

It is often useful to see phenotypic changes with respect to time. For this reason, a scatter plot can be generated that plots these changes for each experimental group. To run this code, you must have a column entitled "**Group**" which contains the groups you wish to compare.

```{r Echo, echo=FALSE}
library(ggplot2)
library(lubridate)
library(tidyr)
#Plot EF over time
All_Echoes<-Annotated_echo %>% drop_na(Sample_ID, Group, `LV Trace_M-Mode_Fractional Shortening`) # Remove NA's
ggplot(data = All_Echoes, aes(x = Duration, y = `LV Trace_M-Mode_Fractional Shortening`, color = Group)) + 
  geom_point() + 
  geom_smooth() + #Need more than 1 date to use this...
  ylim(5,95) +
  theme_bw()
```

# Group-wise comparison

In addition to time-course studies, it can be helpful to compute statistical comparison between experimental groups at each time-point. The following script accomplishes this by iteratively performing pairwise statistics with respect to the "**WT-Sham**" value in the "**Group**" Column.

```{r groupwise}
library(ggplot2)
library(gridExtra)
library(ggpubr)
library(dplyr)
library(gtools)
library(openxlsx)
library(tidyr)
ifelse(!dir.exists(file.path("Echo_Comparisons")), dir.create(file.path("Echo_Comparisons")), FALSE)
COMPARISONS = list(c("WT_Sham", "WT_ORAB"), c("WT_Sham", "BAP1_WT"), c("WT_Sham", "BAP1_ORAB"), c("BAP1_Sham", "BAP1_ORAB"))
## For loop creating a graph for each gene
plotlist = list()
p<-1
All_Echoes<-Annotated_echo %>% drop_na(Sample_ID, Group, `LV Trace_M-Mode_Ejection Fraction`) #Remove NA's
Echo_dates<-unique(All_Echoes$Dates) #Identify all dates in this file
for (i in seq_along(Echo_dates)){
  Echo_time<-All_Echoes %>% filter(Dates==Echo_dates[i]) %>% mutate(LVEF = `LV Trace_M-Mode_Ejection Fraction`)
g_plot<-ggboxplot(Echo_time, x = "Group", 
          y = "LVEF", 
          fill = "Group",
          add = "jitter"
          ) + 
  ggtitle(paste0("LVEF at ", Echo_time$Duration[i], " days")) +
  stat_compare_means(aes(group = Group),
                    comparisons = COMPARISONS, 
                    label = "p.signif"
                    ) +
  theme(axis.text.x=element_text(size=rel(0.75)), axis.text.y=element_text(size=rel(0.75)), legend.position="bottom") + # resize labels, remove legend
  scale_y_continuous(limits = c(0, 100)) # + # expand = expansion(mult = c(0, 0.1)) ### Y scale (to see the statistics)
  # geom_text_repel(aes(label=Sample_ID), , color = "grey", size = 2) # Add Sample_ID
g_plot
pdf(file=paste0("Echo_Comparisons/", Echo_time$Duration[i], "_Echoes.pdf"), width = 6, height = 3)
print(g_plot)
dev.off()
plotlist[[i]] = g_plot
}
t<-marrangeGrob(grobs = plotlist, legend, nrow=2, ncol=1)
ggsave(paste0("Echo_Comparisons/",i, "_ALL.pdf"), t, width = 6.5, height = 9)
t
```

# Survival Plots

```{r}
library(survminer)
library(survminer)
library(RTCGA.clinical)
survivalTCGA(BRCA.clinical, OV.clinical,
             extract.cols = "admin.disease_code") -> BRCAOV.survInfo
library(survival)
fit <- survfit(Surv(times, patient.vital_status) ~ admin.disease_code,
               data = BRCAOV.survInfo)
# Visualize with survminer
ggsurvplot(fit, data = BRCAOV.survInfo, risk.table = TRUE)
###################################################
ggsurvplot(
   fit,                     # survfit object with calculated statistics.
   data = BRCAOV.survInfo,  # data used to fit survival curves. 
   risk.table = TRUE,       # show risk table.
   pval = TRUE,             # show p-value of log-rank test.
   conf.int = TRUE,         # show confidence intervals for 
                            # point estimaes of survival curves.
   xlim = c(0,2000),        # present narrower X axis, but not affect
                            # survival estimates.
   break.time.by = 250,     # break X axis in time intervals by 500.
   ggtheme = theme_minimal(), # customize plot and risk table with a theme.
 risk.table.y.text.col = T, # colour risk table text annotations.
  risk.table.y.text = FALSE # show bars instead of names in text annotations
                            # in legend of risk table
)
```

# **Script References**

```{r references}
Echo_tE<-Sys.time()
Time.Run<-Echo_tE-Echo_tS
Time.Run
options(kableExtra.latex.load_packages = FALSE)
library(kableExtra)
sinfo<-devtools::session_info()
sinfo$platform
sinfo$packages %>% kable( 
                         align="c", 
                         longtable=T, 
                         booktabs=T,
                         caption="Packages and Required Dependencies") %>% 
    kable_styling(latex_options=c("striped", "repeat_header", "condensed"))
```
